---
title: "EDX MovieLens Capstone Project"
author: Mark145
date: Summer 2019
output: 
  pdf_document:
   # toc: true
    #number_selections: true
fontsize: 12pt
geometry: margin=1in
---
```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

#introduction/overview/executive summary
an introduction/overview/executive summary section that describes the dataset and summarizes the goal of the project and key steps that were performed

Do you want insight into how internet based video services 'tailor' movie reccomendations specifically to you? Or do you want to know which aspects of your user data is being used to generate these recommendations?

Here you will be able to view the type of data used by video services and observe machine leaning techniques used to generate movie recommendations.



#methods/analysis
a methods/analysis section that explains the process and techniques used, such as data cleaning, data exploration and visualization, any insights gained, and your modeling approach
a results section that presents the modeling results and discusses the model performance

##aquire the data

The following script will generate 2 tables for us to analyze and model a video services machine learning algorithm against.  The two tables, edx set and validation set, are considered our train and test sets respectivly.

```{r, Create edx set, validation set}
################################
# Create edx set, validation set
################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data

set.seed(1, sample.kind="Rounding")
# if using R 3.5 or earlier, use `set.seed(1)` instead
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

The data aquired from the following script is derived from the MovieLens 10M dataset.  The GroupLens stated summary of the data set is as follows:

"This data set contains 10000054 ratings and 95580 tags applied to 10681 movies by 71567 users of the online movie recommender service MovieLens.

Users were selected at random for inclusion. All users selected had rated at least 20 movies. Unlike previous MovieLens data sets, no demographic information is included. Each user is represented by an id, and no other information is provided." - http://files.grouplens.org/

##explore the data

```{r, set dims}
str(edx)
str(validation)
```

By looking at each set's structure, we see a slight difference in the dimensions.  'edx' has 9000055 objects where as 'validation' only has 999999 objects.  Otherwise the data is setup in the same format, with varing values between the sets.  Below are some visual representations of the data.

```{r}
edx %>% ggplot(aes())
```


#conclusion
a conclusion section that gives a brief summary of the report, its limitations and future work (the last two are recommended but not necessary)